Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-21 submission/manuscript.Rmd
results edit 4

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Therefore, we wondered if _Kdm5c_ mutant males express sperm genes while mutant females express egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes. We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H). We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J). Germline genes dysregulated exclusively in_Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex. This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs. These results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.

>>>>

While many germline genes act in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Therefore, we wondered if _Kdm5c_ mutant males were more likely to express sperm genes and if mutant females would instead express egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we filtered our list of germline-enriched genes for egg and sperm-biased genes. We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes (Figure 4H). We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants (Figure 4I-J). Germline genes dysregulated exclusively in _Kdm5c_ mutant males and females were also not biased towards their corresponding germ cell sex (Figure 4I). This indicates sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs. Altogether, these results demonstrate that the sex of _Kdm5c_ mutant cells influences the degree of germline gene, independent of germline gene sex.
